Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-09-10 review/response.Rmd

### Major comment

-   "Sometimes, two optimisers can start with the same basis but finish with bases of opposite sign...However, this creates difficulties for comparing the optimisers since the end bases will be symmetric to the origin. A sign flipping step is conducted..." and the "Reconciling the orientation" section

    -   For 1D projections, flipping the sign is sufficient. For p-D projections of $\mathbb{R}^n$​, one may need a little more. To my knowledge most index functions are invariant to any rotations that happens within the p-subspace. This makes me wonder that in general, one may want to align two bases so that their difference does not contain any "within-subspace" components. In other words, one may want to compare representations in the Grassmannian manifold (space of k-subspaces) instead of in the Stiefel manifold (space of orthonormal k-frames).

\textcolor{red}{We agree that most index functions should be rotation invariant (this is in fact a desirable property, but there are exceptions where indexes can still be very useful but not rotation invariant, see discussion in Laa, Cook (2020)). Thus when comparing frames defining projections to d>=2 dimensions, we may want to introduce a rotation that will align the bases as much as possible within the common subspace, while potentially quantifying e.g. the angle between two planes. This is an interesting problem, however the implementation of this general approach is beyond the scope of the current paper.}

@@

### Minor comment

-   The software seems to be specific to projection pursuit. I wish the authors could comment more on its potential generality to other optimizations. For example, can one generalizes the tour method to visualize any parameter trace of an optimizer?

\textcolor{red}{It is feasible to generalise the tour method for diagnosing other optimisers, given that the parameter space can be simulated. In our examples, the orthonormality constraint regulates the parameter space of the projection basis ($p \times d$) to be an p-sphere, which can be simulated by the geozoo package. For other optimisers, reserachers need first to create a data collection object in a tibble that tracks the parameter during the optimisation, and then to use ferrn to produce the diagnostic plots as shown in the paper. Some examples of the data collection objects that we created are included in the ferrn package and should be referred to as the format for inputs into the expore\_\* functions.}

@@

\textcolor{red}{Thanks for sharing your interpretation on the effects of flipping, that's right! The reviewer is giving a decent explanation in describing the effect of flipping. The gray dot should land on the dashed line and this is a mistake due to the data not being updated after flipping in the code. This has now been corrected.}

-   "Recall that the columns in a 2D basis are orthogonal to each other, so the space of *p* × 2 bases is a *p*-D torus".

    -   It is not clear to me why "so". For p=2, the space of 2x2 bases is a pair of 1-D tori instead of just one (because you have two orientations for a pair of orthogonal vectors), right? For p=3, the space of 3x2 bases is SO(3), which is not a 3-D torus? Ref: <https://en.wikipedia.org/wiki/Stiefel_manifold>

\textcolor{red}{We changed the wording of the sentence to: "a torus in the $p$-D space" and provide a reference for this statement: (Buja and Asimov, 1986).}

### 2021-09-10 paper/zhang-cook-laa-langrene-menendez.Rmd
add reference of torus in p-D space and more explanation on flipping

While the previous few examples have looked at the space of 1-frames in a unit sphere, this section visualises the space of 2-frames. Recall that the columns in a 2D basis are orthogonal to each other, so the space of $p \times 2$ bases is a torus in the $p$-D space. For $p = 3$ one would see a classical 3D torus shape as shown by the grey points in Figure \ref{fig:toy-torus}. The two circles of the torus can be observed to be perpendicular to each other and this can be linked back to the orthogonality condition. Two paths from CRS and PD are plotted on top of the torus and coloured in green and brown respectively to match the previous plots. The final basis found by PD and CRS are shown in a larger shape and printed below respectively:

@@

One interesting situation observed in the previous examples is that, for some simulations, as shown on the left panel of Figure \ref{fig:flip-sign}, the target basis is generated on the other half of the basis space and the interpolator seems to draw a straight line to interpolate. As mentioned previously, bases with opposite signs do not affect the projection and index value, but clearly, we prefer the target to have the same orientation as the current basis. The orientation of two bases can be checked via calculating the determinant and a negative value indicates opposite direction. Hence an orientation check is carried out once a new target basis is generated and the sign in the target basis will be flipped if a negative determinant is obtained. The interpolation after implementing the orientation check is shown on the right panel of Figure \ref{fig:flip-sign} where the unsatisfactory interpolation no longer exists.

>>>>

While the previous few examples have looked at the space of 1-frames in a unit sphere, this section visualises the space of 2-frames. Recall that the columns in a 2D basis are orthogonal to each other, so the space of $p \times 2$ bases is a torus in the $p$-D space [@Buja1986-sw]. For $p = 3$ one would see a classical 3D torus shape as shown by the grey points in Figure \ref{fig:toy-torus}. The two circles of the torus can be observed to be perpendicular to each other and this can be linked back to the orthogonality condition. Two paths from CRS and PD are plotted on top of the torus and coloured in green and brown respectively to match the previous plots. The final basis found by PD and CRS are shown in a larger shape and printed below respectively:

@@

One interesting situation observed in the previous examples is that, for some simulations, as shown on the left panel of Figure \ref{fig:flip-sign}, the target basis is generated on the other half of the basis space and the interpolator seems to draw a straight line to interpolate. Bases with opposite signs do not affect the projection and index value, but we would prefer the target to have the same orientation as the current basis. The orientation of two bases can be checked via calculating the determinant and a negative value suggests the two bases having different orientation. For 1D bases, this can be corrected by flipping the sign in one of the basis while the correction for a general p-D bases is more complex and beyond the scope of this paper. Here an orientation check is carried out once a new target basis is generated and the sign in the target basis will be flipped if a negative determinant is obtained. The interpolation after implementing the orientation check is shown on the right panel of Figure \ref{fig:flip-sign} where the unsatisfactory interpolation no longer exists.

### 2021-09-10 paper/zhang-cook-laa-langrene-menendez.tex
add reference of torus in p-D space and more explanation on flipping

While the previous few examples have looked at the space of 1-frames in
a unit sphere, this section visualises the space of 2-frames. Recall
that the columns in a 2D basis are orthogonal to each other, so the
space of \(p \times 2\) bases is a torus in the \(p\)-D space. For
\(p = 3\) one would see a classical 3D torus shape as shown by the grey
points in Figure \ref{fig:toy-torus}. The two circles of the torus can
be observed to be perpendicular to each other and this can be linked
back to the orthogonality condition. Two paths from CRS and PD are
plotted on top of the torus and coloured in green and brown respectively
to match the previous plots. The final basis found by PD and CRS are
shown in a larger shape and printed below respectively:

@@

Experiments in this paper have shown that random search algorithms tend
to require larger search size for 2D indexes as compared to 1D indexes.
All the three optimisers tend to work well for smooth indexes while PD
generally fails for non-smooth indexes. This is due to the fact that
derivative based methods are in general not suitable for optimising
non-smooth functions.

For other researchers, the tools implemented in the ferrn package can
track the progression of the target function and monitor parameter
coverage of the search space in both reduced 2D space and the full
space. These visualisation can lead to insights for evaluating the
performance of multiple optimisers in a specific problem, as the example
shown for comparing the three optimisers when optimising the kolmogorov
index in the paper.

>>>>

While the previous few examples have looked at the space of 1-frames in
a unit sphere, this section visualises the space of 2-frames. Recall
that the columns in a 2D basis are orthogonal to each other, so the
space of \(p \times 2\) bases is a torus in the \(p\)-D space
\citep{Buja1986-sw}. For \(p = 3\) one would see a classical 3D torus
shape as shown by the grey points in Figure \ref{fig:toy-torus}. The two
circles of the torus can be observed to be perpendicular to each other
and this can be linked back to the orthogonality condition. Two paths
from CRS and PD are plotted on top of the torus and coloured in green
and brown respectively to match the previous plots. The final basis
found by PD and CRS are shown in a larger shape and printed below
respectively:

@@

The tools were implemented in the ferrn package and can track the
progression of the target function and monitor parameter coverage of the
search space in both reduced 2D space and the full space. These
visualisation can lead to insights for evaluating and comparing the
performance of multiple optimisers in a specific problem. They can
provide better understanding of existing methods or motivate the
development of new approaches. For example, we have compared how three
optimisers perform when maximising a non-smooth index function and
illustrated how the pseudo-derivative search fails in this setting. The
observations from our experiments have also been translated into
improved optimisation methods for the guided tour, e.g.~we introduced
the option to interrupt the search if a better basis is found along the
path.
